Best Practices in Mixture Modeling using Free Open Source Software

### 2022-08-31 manuscript.Rmd
Updates 23 August
Updated the manuscript.

All the references are now in the [@key] format.  New .bib file  with references has been made.

Reworked the model fit section to reflect statistics tidySEM reports.
Stated that LRT cannot be used.
Referred to the bootstrapped LRT being available albeit extremely slow; we are currently developing a faster version.

Adjusted classification diagnostics to reflect what tidySEM provides. Entropy is finished. Needs more information on the range of the posterior classification probabilities by most likely class membership and the range of the sample size.

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the OpenMx R-package.

@@

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibsonTenFrequentlyAsked2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that latent class analysis 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
*unsupervised classification* problem [@figueiredoUnsupervisedLearningFinite2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).

@@

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermuntj.k.LatentClassAnalysis2004].
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

@@

<!-- f a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

<!-- ecause LPA is model-based, a number of different model parameterizations can be -->
<!-- estimated. These models differ in terms of whether--and how--parameters are -->
<!-- estimated across the profiles. These parameters are the means for the different -->
<!-- profiles, which, in this approach, always are estimated freely across the -->
<!-- profiles; the variances for the variables used to create the profiles, which can -->
<!-- be estimated freely or can be estimated to be the same, or equal, across -->
<!-- profiles; and the covariances of the variables used to create the profiles, -->
<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->

@@

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughmanMixtureModelAnalysis2006] or mental [@wuAbuseDependencePrescription2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

@@

<!--                                                1 -->
<!-- A -->
<!-- downside is that these models are more complex (much more parameters to be estimated). -->
<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->

@@

The best practices in estimation, as outlined in Table \@ref(tab:checkest),
are rooted in existing recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis [@vandeschootGRoLTSChecklistGuidelinesReporting2017]
and latent class analysis with ordinal indicators [e.g., @nylund-gibsonTenFrequentlyAsked2018].
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

@@

Examining observed data is essential for any analysis,
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermuntKmeansMayPerform2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @normanLikertScalesLevels2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [REF SPURK].
This is incorrect.
When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
<!-- CJ: Expand on this here -->

Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @vanlissaWORCSWorkflowOpen2020].

@@

Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@schootGRoLTSChecklistGuidelinesReporting2017].
Three types of missingness have been distinguished in the literature [@rubinInferenceMissingData1976]:
Missing completely at random (MCAR), which means that missingness is random;
missing at random (MAR), which means that missingness is contingent on the *observed* data (and can thus be accounted for);
and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
It is possible to conduct a so-called "MCAR" test, 
for example the non-parametric MCAR test [@jamshidianTestsHomoscedasticityNormality2010].
But note that the name "MCAR test" is somewhat misleading,
as the null-hypothesis of this test is that the data are not MAR,
and a significant test statistic indicates that missingness is related to the observed data (MAR).
A non-significant test statistic does not distinguish between MCAR or MNAR.
As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@littleTestMissingCompletely1988].
This assumption is tenuous in the context of latent class analysis.
A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [-@jamshidianTestsHomoscedasticityNormality2010].
Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
For this tutorial, I have re-implemented it in the `mice` package as `mice::mcar()`,
with a fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due dilligence,
it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; on par with multiple imputation [@leeComparisonFullInformation2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.

@@

The different types of latent class models have different parameters.
For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
Latent growth analyses have the same parameters, but with respect to the latent growth variables.
Latent class analyses with ordinal indicators have thresholds.
All of these parameters can be freely estimated across classes, or constrained,
or fixed (e.g., to zero).
The total number of parameters thus scales with the number of estimated classes.
Consequently, latent class analyses have a potentially very high number of parameters.
As any of these parameters could be mis-specified,
it is important to consider alternative model specifications.
However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
This distinction has remained underemphasized in prior writing.

Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylundDecidingNumberClasses2007].
In exploratory analyses, a large number of models are typically estimated in batch,
with varying numbers of classes and model specifications.
The "correct" model specification is then determined based on a combination of fit indices,
significance tests,
and interpretability.
For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
largely automates this process.
The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
The means are free to vary across classes by default, although the more general function `tidySEM::mx_mixture()` could be used to circumvent this.
After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
because this test is very computationally expensive.
It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
Fit indices typically used for determining the optimal number of classes include
the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
and add a penalty for the number of parameters (thus incentivizing simpler models).
This helps balance fit and model complexity.
The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.

Fit indices may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
One option is to select a specific fit index before analyzing the data.
Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogulComparisonInformationCriteria2016].

Confirmatory analyses typically require less comprehensive alternative model specifications.
For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.

Van de Schoot and colleagues [-@vandeschootGRoLTSChecklistGuidelinesReporting2017] argue that there are many choices to be made in the specification of latent trajectory models,
and that alternative

@@

Conversely, the free open source program OpenMx is very flexible,
but not very user-friendly.

New functionality in the R-package `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
It adheres to best practices in estimation and reporting, as described in this paper.
The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
This syntax offers a human-readable way to specify latent variable models.
Minor enhancements are made to simplify the specification of latent class analysis.

Because of the limitations in ex tools, we set out to develop a tool that a)
provided sensible defaults and were easy to use, but provided the option to
access and modify all of the inputs to the model (i.e., low barrier, high
ceiling), b) interfaced to existing tools, and are able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify,

fully-reproducible analyses and

<!-- 3. Packages -->
<!-- 3.1 R-packages -->
<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->

@@

Mixture model parameters and model fit statistics can be estimated in a variety of ways.
The choice of the estimator depends on the presence of missing values,
sample size, number of indicators, and available computational resources (Weller, Bowen & Faubert, 2020).
A commonly used technique is maximum likelihood (ML) estimation
with the expectation-maximization (EM) algorithm as a local optimizer. 
Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
We can think of this optimization problem as a three-dimensional landscape: 
The X and Y dimensions are determined by the class-specific means, 
so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
The optimizer must find the deepest "valley" in this landscape, 
which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
The EM optimizer behaves somewhat like a marble, dropped in this landscape.
It is dropped at some random point in space, and will roll into the nearest valley.
The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
(this is known as "convergence").
It cannot climb out again.
Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
This is the "random starts" approach.

@@

One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurkLatentProfileAnalysis2020]. 
This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
<!-- CJ: Is there even any literature that says robust standard erorrs are better with latent class analysis? -->

@@

LCA can be done in an exploratory or in a confirmatory fashion.
In exploratory LCA, a sequence of models is fitted to the data 
with each additional model estimating one more class than the previous model.
These models are then compared and the best solution is selected as the final class solution.
In some cases, prior theory can inform the researcher about the number of classes to expect.
Even in such confirmatory LCA cases, it is nonetheless useful to know 
if the theoretical model is markedly better than those with differing numbers of classes.

>>>>

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the OpenMx R-package.

@@

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that latent class analysis 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).

@@

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermunt_jk_latent_2004].
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

@@

<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

<!-- Because LPA is model-based, a number of different model parameterizations can be -->
<!-- estimated. These models differ in terms of whether--and how--parameters are -->
<!-- estimated across the profiles. These parameters are the means for the different -->
<!-- profiles, which, in this approach, always are estimated freely across the -->
<!-- profiles; the variances for the variables used to create the profiles, which can -->
<!-- be estimated freely or can be estimated to be the same, or equal, across -->
<!-- profiles; and the covariances of the variables used to create the profiles, -->
<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->

@@

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

@@

<!--                                                1 -->
<!-- A downside is that these models are more complex (much more parameters to be estimated). -->
<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->

@@

The best practices in estimation, as outlined in Table \@ref(tab:checkest),
are rooted in existing recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

@@

Examining observed data is essential for any analysis,
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
<!-- CJ: Expand on this here -->

Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021.

<!--  Adjust the following part, it belongs here:  -->

<!-- CJ: This is about model interpretation; conditional item probabilities are the analogue of mean and standard deviation, but when the observed data are binary/ordinal  -->
<!-- Conditional item probabilities indicate the probability of a positive score on an item  -->
<!-- given that the person belongs to a particular latent class.  -->

<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
the classes have high homogeneity with respect to this item [@masyn_latent_2013]. -->
<!-- However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),  -->
<!-- we are observing boundary parameter estimates. This can be a sign of an invalid solution,  -->
<!-- warn us that too many classes were extracted, or indicate a local optimum.  -->
<!-- Boundary parameter estimates are more likely to happen when working with sparse data  -->
<!-- and should be interpreted with caution (Geiser, 2012).  -->

<!-- CJ: This talks about the problem of inadmissible solutions. With ordinal items, that can happen because of sparse cross-tables (e.g., some combinations of responses never occur). WIth continuous items, that could happen e.g. because you have a zero-inflated, otherwise normally distributed variable; one class would have a mean at zero and its SD would be hard to determine. Or it could happen if you model a five-category Likert scale with a five-component latent profile analysis; each response category will have its own mean, but there's no variance around that (everybody in class 3 has exactly the response value 3). These are all the same problems. -->

<!-- An important outcome of LCA are unconditional class probabilities.  -->
<!-- These communicate the probability that an individual belongs to each class.  -->
<!-- A limitation is that our final class solution need not reflect true class membership.  -->
<!-- Once we do accept a particular class solution, special care also needs to be taken when naming the classes.  -->

A problem which can occur is that of inadmissible solutions. 
With binary indicators, LCA is modelling a cross-table with all the predictors.
The problem with such cross-tables is that they will often contain empty cells,
i.e. combinations of responses that never occur together. 
This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
Such boundary parameter estimates could indicate that the solution is invalid.
Boundary parameter estimates can also happen with continuous indicators.
For instance, if we have a zero-inflated normal distribution and a two class solution,
one class might have the mean of zero and its standard deviation cannot be determined 
since there is little variance. 
This too could be a sign of an invalid solution,
warn us that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].

@@

Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@van_de_schoot_grolts-checklist_2017].
Three types of missingness have been distinguished in the literature [@rubin_inference_1976]:
Missing completely at random (MCAR), which means that missingness is random;
missing at random (MAR), which means that missingness is contingent on the *observed* data (and can thus be accounted for);
and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
It is possible to conduct a so-called "MCAR" test, 
for example the non-parametric MCAR test [@jamshidian_tests_2010].
But note that the name "MCAR test" is somewhat misleading,
as the null-hypothesis of this test is that the data are not MAR,
and a significant test statistic indicates that missingness is related to the observed data (MAR).
A non-significant test statistic does not distinguish between MCAR or MNAR.
As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
This assumption is tenuous in the context of latent class analysis.
A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [@jamshidian_tests_2010].
Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
For this tutorial, I have re-implemented it in the `mice` package as `mice::mcar()`,
with a fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due dilligence,
it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; on par with multiple imputation [@lee_comparison_2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.

@@

The different types of latent class models have different parameters.
For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
Latent growth analyses have the same parameters, but with respect to the latent growth variables.
Latent class analyses with ordinal indicators have thresholds.
All of these parameters can be freely estimated across classes, constrained to be equal across classes,
or fixed to a certain value (e.g., to zero).
The total number of parameters thus scales with the number of estimated classes.
Consequently, latent class analyses have a potentially very high number of parameters.
As any of these parameters could be misspecified,
it is important to consider alternative model specifications.
However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
This distinction has remained underemphasized in prior writing.

Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylund_deciding_2007].
In exploratory analyses, a large number of models are typically estimated in batch,
with varying numbers of classes and model specifications.
The "correct" model specification is then determined based on a combination of fit indices,
significance tests,
and interpretability.
For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
largely automates this process.
The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
The means are free to vary across classes by default, although the more general function `tidySEM::mx_mixture()` could be used to circumvent this.
After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
because this test is very computationally expensive.
It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices. More on fit indices can be found in the Model fit indices subsection of this paper.

Confirmatory analyses typically require less comprehensive alternative model specifications.
For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.
Therefore, it may not be necessary to test different model specifications when using the confirmatory approach.

<!-- see the following article and expand on the paragraph -->
@van_de_schoot_grolts-checklist_2017 argue that there are many choices to be made in the specification of latent trajectory models,
and that alternative

@@

Conversely, the free open source program OpenMx is very flexible,
but not very user-friendly. We directly address this limitation using the `tidySEM` R-package.
New functionality in `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
It adheres to best practices in estimation and reporting, as described in this paper.
The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
This syntax offers a human-readable way to specify latent variable models.
Minor enhancements are made to simplify the specification of latent class analysis.

Because of the limitations of the aforementioned tools, we set out to develop a free tool that
provides sensible defaults and is easy to use, but provides the option to
access and modify all of the model inputs (i.e., low barrier, high
ceiling). `tidySEM` interfaces with existing tools, and is able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify. 
Furthermore,`tidySEM` facilitates fully-reproducible analyses and contributes to open science.

<!-- 3. Packages -->
<!-- 3.1 R-packages -->
<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->

@@

Mixture model parameters and model fit statistics can be estimated in a variety of ways.
The choice of the estimator depends on the presence of missing values,
sample size, number of indicators, and available computational resources [@weller_latent_2020].
A commonly used technique is maximum likelihood (ML) estimation
with the expectation-maximization (EM) algorithm as a local optimizer. 
Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
We can think of this optimization problem as a three-dimensional landscape: 
The X and Y dimensions are determined by the class-specific means, 
so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
The optimizer must find the deepest "valley" in this landscape, 
which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
The EM optimizer behaves somewhat like a marble, dropped in this landscape.
It is dropped at some random point in space, and will roll into the nearest valley.
The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
(this is known as "convergence").
It cannot climb out again.
Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
This is the "random starts" approach.

@@

One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->

@@

LCA can be done in an exploratory or in a confirmatory fashion.
In exploratory LCA, a sequence of models is fitted to the data 
with each additional model estimating one more class than the previous model.
These models are then compared and the best solution is selected as the final class solution.
In some cases, prior theory can inform the researcher about the number of classes to expect.
Even in such confirmatory LCA cases, it is nonetheless useful to know 
if the theoretical model is markedly better than those with differing numbers of classes.
Therefore, it may always be useful to compare different class solutions.
